---
title: "What Did McCleary Buy?"
subtitle: "A synthetic-control estimate of Washington's school finance reform"
author: "Richard Sprague"
date: today
---

```{r setup}
#| include: false
source(here::here("R", "00_config.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "04_build_panel.R"))
source(here::here("R", "05_synth.R"))
source(here::here("R", "06_plots.R"))

panel <- build_analysis_panel()
fits  <- run_all(panel)
inf   <- synth_inference(fits$score)

## Washington 2011 -> 2019 growth as the F-33 pipeline computes it, for the
## comparison against the OSPI figures quoted in "The question".
f33 <- local({
  d <- dplyr::filter(panel, state == TREAT_UNIT, year %in% c(2011, 2019))
  pc <- function(col) 100 * (d[[col]][2] / d[[col]][1] - 1)
  list(
    total      = pc("rev_pp"),
    state      = pc("rev_state_pp"),
    local      = pc("rev_local_pp"),
    share_2011 = 100 * d$state_share[1],
    share_2019 = 100 * d$state_share[2]
  )
})
```

## The constitutional promise

Washington's constitution opens its education article with 34 words that have
shaped the state's education policy ever since [@bazzaz2026paramount]:

> It is the paramount duty of the state to make ample provision for the
> education of all children residing within its borders without distinction or
> preference on account of race, color, caste, or sex.

*Paramount duty* is a legal standard, not a slogan, and since the 1970s
Washington's courts have repeatedly found the state in breach of it. The most
recent finding was **McCleary v. Washington** (2012), brought by a coalition of
families, school districts, and the statewide teachers union. The state Supreme
Court found the underfunding so severe that it held the legislature in contempt
and **fined the state $100,000 a day for three years** until lawmakers
appropriated the money.

The Legislature appropriated it, and K–12 became the largest single item in
the state budget, buying smaller class sizes, some of the highest teacher pay
in the country, and per-student spending just above the national average.

### What the money bought

Bazzaz's survey of the results is careful and, on achievement, muted.
Washington performs slightly ahead of the nation in NAEP reading and roughly at
the national average in math, with scores below pre-pandemic levels here as
everywhere. The state's own goal — 90% of students proficient on its state
assessments — sits well above the roughly 70% at grade level in reading and
two-thirds in math. Underneath the averages are sharp disparities: only 42% of
Washington's Latino fourth graders reach NAEP's "basic" level in reading.

State superintendent Chris Reykdal frames the constitutional money as a floor
rather than a program, and puts the open question plainly:

> Money really does matter. But how you deploy it matters too.

That is the question this project takes up. The Seattle Times piece is the
first in a series that promises to examine "what results the state gets — and
doesn't get — for its substantial investment." **This repo is an attempt to
answer one narrow, quantitative slice of that question with an explicit
counterfactual rather than a before-and-after comparison.**

The narrowing matters. This analysis can say
something about grade 8 mathematics, because NAEP provides a comparable
measure across states. It cannot evaluate the two areas where Bazzaz finds
Washington genuinely lagging — preschool access, where the state ranks 7th in
per-child spending but 34th in enrollment, and postsecondary completion, where
its six-year rate of 56.5% is among the nation's worst. Both fall outside the
constitutional guarantee, and neither has outcome data that supports this
design.

## What NAEP measures

The **National Assessment of Educational Progress** — "The Nation's Report
Card" — is a federally administered test given to a *representative sample* of
students in every state, not to every student [@naep2024mathematics]. It is run
by the National Center for Education Statistics in grades 4 and 8, in
mathematics and reading, in odd-numbered years (the pandemic pushed the 2021
administration to 2022).

Three properties make it the only usable outcome here:

1. **It is the same test everywhere.** States write their own standards, their
   own assessments, and their own definitions of "proficient." Washington's
   goal of 90% proficiency is measured on Washington's exam, which tells you
   nothing about Oregon. NAEP is a common instrument, which is what makes the
   cross-state comparison in this analysis possible at all.
2. **No state can move its own score by redefining the standard.** Because NAEP
   is not tied to any state's curriculum or cut scores, it is largely immune to
   the standards-inflation that makes state proficiency rates incomparable over
   time.
3. **It is reported as a mean scale score**, on a 0–500 scale. This analysis
   uses the mean rather than the percent above a proficiency threshold, because
   a threshold discards everything about where students sit within a band, and
   moves for reasons that have nothing to do with the middle of the
   distribution.

The outcome here is **grade 8 mathematics**. Grade 8 has the longest clean
run of state-level results, and mathematics is more responsive to schooling —
and less to home environment — than reading, which makes it the more sensible
place to look for an effect of a school-funding change. Scale scores move
slowly: differences of two or three points between states are small, and the
gap the analysis is trying to detect is of that order.

## The question

Between the 2009–11 and 2019–21 biennia, Washington's state spending on K–12
education rose 110%, while all other state spending rose 52% [@wrc2020mccleary].
Over the same period the state's standing on the National Assessment of
Educational Progress did not improve relative to the rest of the country. In
grade 8 mathematics Washington led the national public average by 5 points in
2003 and by 2 points in 2024.

Those two facts are usually deployed as a rhetorical pair. They should not be.
Three things have to be settled before they can be compared at all:

1. **How much money actually reached schools per student?** The 110% figure
   measures the state budget line. McCleary's central mechanism was a levy
   swap — the state assumed obligations previously met by local levies — so
   some share of that growth is a change in *who writes the check*, not in how
   much is written. Using OSPI's district accounting, the Washington Research
   Council reports that from SY 2010–11 to SY 2018–19 total per-pupil revenues
   rose 49.7%, state per-pupil revenues rose 81.6%, and local taxes per pupil
   fell 0.3% [@wrc2020mccleary].

   The pipeline behind this document does not reproduce that last figure. It
   uses the Census F-33 survey rather than OSPI's F-196, and over the same
   years F-33 puts Washington's local revenue per pupil
   **up `r sprintf("%.1f%%", f33$local)`**, with local property tax alone up
   14.9% — not flat, and not falling. The two sources agree closely on the
   totals (F-33 gives `r sprintf("%+.1f%%", f33$total)` total and
   `r sprintf("%+.1f%%", f33$state)` state) and on the direction of the levy
   swap: state share of revenue rises about
   `r round(f33$share_2019 - f33$share_2011)` points in F-33
   (`r sprintf("%.1f%%", f33$share_2011)` to `r sprintf("%.1f%%", f33$share_2019)`)
   and about 14 in OSPI (64.6% to 78.4%). They disagree on local revenue in
   sign, not just magnitude. The
   disagreement is not a definitional one about which local receipts count, and
   not a year-alignment artifact; both were checked. It is a genuine difference
   between two federal and state collections, and it is left visible here
   rather than resolved by picking the more convenient number. Nothing in the
   estimates below depends on it: the outcome model uses NAEP scores only, and
   the first stage uses total revenue, where the sources agree.

2. **What were those dollars worth?** Nominal per-pupil growth has to be
   deflated twice: once for inflation, once for the fact that Puget Sound labor
   got more expensive relative to the rest of the country over the same decade.
3. **Compared to what?** Washington's post-2019 scores fell. So did everyone's.
   The counterfactual is not "Washington in 2011"; it is "Washington had the
   legislature not responded to McCleary."

This project answers the third question with a synthetic control, and carries
the first two through the data pipeline so the resource measure being tested is
one that a skeptic would accept.

## Data

```{r}
#| label: tbl-panel
#| tbl-cap: "Analysis panel coverage"
panel |>
  dplyr::summarise(
    States = dplyr::n_distinct(state),
    `NAEP years` = dplyr::n_distinct(year),
    `First year` = min(year),
    `Last year` = max(year),
    Observations = dplyr::n()
  )
```

Three sources, all public, all pulled by script:

| Source | Series | Access |
|---|---|---|
| NAEP Data Service (NCES) | State mean scale scores, grade 8 math | `R/01_fetch_naep.R` |
| Census F-33 / CCD via Urban Institute | Revenue and expenditure per pupil | `R/02_fetch_finance.R` |
| BEA Regional Price Parities; CPI-U | Place and time deflators | `R/03_price_adjust.R` |

Every API pull is validated against hand-transcribed published figures in
`data/cache/naep_verified_anchors.csv` before the analysis runs
(`make test`). This guards against the failure mode where an upstream schema
change silently returns a different series and the analysis keeps running.

### The spending series, deflated

```{r}
#| label: fig-spending
#| fig-cap: "Washington per-pupil revenue, nominal and deflated"
plot_spending(panel)
```

```{r}
#| label: tbl-spending
#| tbl-cap: "Washington per-pupil revenue by NAEP year"
panel |>
  dplyr::filter(state == TREAT_UNIT) |>
  dplyr::transmute(
    Year = year,
    `Nominal $/pupil` = round(rev_pp),
    `Real, cost-adj. $/pupil` = round(rev_pp_real),
    `State share of revenue` = scales::percent(state_share, accuracy = 0.1)
  )
```

## Design

### One state, no control group

The obvious way to evaluate McCleary is to compare Washington before and after.
That fails immediately, because 2022 and 2024 scores fell across the entire
country. Any before-and-after comparison credits McCleary with a pandemic.

The next idea is to pick a comparison state. But no single state is a good
match for Washington on demographics, economy, and prior test-score trajectory
at once, and whichever one you pick, a skeptic can reasonably accuse you of
having picked it to get the answer you wanted.

**Synthetic control** [@abadie2010synthetic] solves both problems by refusing
to choose. Instead of one comparison state, it builds a weighted average of
many, with the weights chosen by an explicit rule rather than by the analyst.

### What synthetic Washington is

Synthetic Washington is a **weighted average of other states' NAEP scores**,
where the weights are picked so that the weighted average tracks Washington's
actual scores as closely as possible *in the years before 2013*.

Two constraints make it honest:

- **Weights cannot be negative**, so the method can never construct a
  counterfactual by subtracting a state.
- **Weights sum to 1**, so synthetic Washington is always an interpolation
  among real states, never an extrapolation beyond the range of anything
  actually observed.

Concretely, the optimizer in this analysis lands on:

```{r}
#| label: tbl-weights-inline
#| tbl-cap: "The states that make up synthetic Washington"
fits$score |>
  tidysynth::grab_unit_weights() |>
  dplyr::filter(weight > 0.001) |>
  dplyr::arrange(dplyr::desc(weight)) |>
  dplyr::transmute(State = unit, Weight = scales::percent(weight, accuracy = 0.1))
```

Read that as: the mix of states whose combined grade 8 math path most closely
reproduced Washington's own path from 2003 through 2011. From 2013 onward, that
same fixed mix is carried forward, and **the difference between Washington and
this synthetic Washington is the estimated effect**.

### How the weights are calculated

The weights come from a constrained optimization, not from judgment. The
procedure minimizes the distance between Washington and the weighted donor pool
across a set of **predictors** — here, Washington's NAEP score in each of the
five pre-treatment years (2003, 2005, 2007, 2009, 2011) — subject to the
non-negativity and sum-to-one constraints above. That is a quadratic program,
solved numerically; `tidysynth` hands it to an interior-point solver.

Note what is *not* in the predictor list: per-pupil revenue. Money is the
treatment mechanism, estimated separately below as a first stage, not something
the donor pool is asked to match on. With only five pre-treatment periods, the
predictor set cannot identify more than five predictors — adding spending terms
on top of the five score lags makes the problem numerically singular. `R/05_synth.R`
documents this at the point of decision.

Treatment is dated to **2013**, the first NAEP administration after the January
2012 ruling. The money phased in through 2019, so the estimated effect is a
dose that grows over the post-period rather than a clean step.

The donor pool excludes states with their own major finance-reform shocks in
2008–2022 — `r paste(EXCLUDE_DONORS, collapse = ", ")` — which leaves
`r length(donor_states(panel))` donors. A state that had its own court-ordered
funding overhaul is not an untreated comparison. That exclusion list is the
most contestable input in this analysis; `notes/methodology.qmd` reports how
much it matters, and the short answer is: almost none.

### Placebo tests

With one treated unit there is no sampling distribution and no standard error
in the usual sense. A gap of four points between Washington and its synthetic
might be an effect, or might be the kind of wobble this method produces
routinely. The only way to know is to find out what "routinely" looks like.

So the procedure is run again, **pretending each donor state was the one
treated in 2013**. Nothing happened to Nebraska in 2013, so whatever gap the
method produces for Nebraska is pure noise — a measure of how much apparent
"effect" this design manufactures when there is nothing to find. Doing this for
every donor gives a distribution of placebo gaps to judge Washington against.

Each unit is scored by its **post/pre RMSPE ratio**. RMSPE is root mean squared
prediction error — the typical size of the gap between a state and its
synthetic. Taking the ratio of the post-treatment value to the pre-treatment
one normalizes for fit quality: a state the donor pool could never match well
to begin with will have large gaps afterward for entirely uninteresting
reasons, and dividing by its pre-period error discounts that.

Rank Washington among those ratios and you get a p-value directly: if
Washington ranks *k*-th of *N*, then *p* = *k* / *N*. It answers the question
"what fraction of states show a post-treatment divergence at least this large
when nothing happened to them?" A Washington that sits unremarkably in the
middle of the placebo pack is a Washington whose gap is indistinguishable from
noise.

## Results

### First stage: did the money actually move?

A finance reform that does not change real resources cannot be tested for
effects on achievement. This is the check that the treatment exists.

```{r}
#| label: fig-first-stage
#| fig-cap: "Real cost-adjusted revenue per pupil: Washington vs synthetic Washington"
plot_trends(
  fits$money,
  title = "Washington did buy something",
  subtitle = "Real, cost-of-living-adjusted revenue per pupil against a synthetic control built from untreated states",
  y_lab = "Revenue per pupil (2019 $, cost-adjusted)"
)
```

### Outcome: NAEP grade 8 mathematics

```{r}
#| label: fig-outcome
#| fig-cap: "NAEP grade 8 math: Washington vs synthetic Washington"
plot_trends(
  fits$score,
  title = "Observed and synthetic Washington",
  subtitle = "Grade 8 mathematics mean scale score"
)
```

```{r}
#| label: fig-gap
#| fig-cap: "Estimated effect: observed minus synthetic"
plot_gap(
  fits$score,
  title = "Estimated effect of the McCleary response on grade 8 math",
  subtitle = "Positive values mean Washington outperformed its synthetic counterfactual"
)
```

### Inference

```{r}
#| label: fig-placebo
#| fig-cap: "Placebo distribution"
plot_placebos(
  fits$score,
  title = "Washington's gap against every donor state's placebo gap",
  subtitle = "If Washington's line is unremarkable inside the gray band, the estimate is indistinguishable from noise"
)
```

```{r}
#| label: tbl-inference
#| tbl-cap: "Post/pre RMSPE ratios, ranked"
inf$table |> head(12)
```

Washington's post/pre RMSPE ratio is `r round(inf$rmspe_ratio, 2)`, ranking
`r inf$rank` of `r nrow(inf$table)` units, giving a placebo-based
*p* = `r round(inf$p_value, 3)`.

## The answer

```{r}
#| label: bound
#| include: false
placebo_gaps <- fits$score |>
  tidysynth::grab_synthetic_control(placebo = TRUE) |>
  dplyr::mutate(gap = real_y - synth_y) |>
  dplyr::filter(time_unit >= TREAT_YEAR, .id != TREAT_UNIT) |>
  dplyr::group_by(.id) |>
  dplyr::summarise(mean_abs_gap = mean(abs(gap)), .groups = "drop")

wa_abs_gap <- synth_effects(fits$score) |>
  dplyr::filter(time_unit >= TREAT_YEAR) |>
  dplyr::summarise(m = mean(abs(gap))) |>
  dplyr::pull(m)

placebo_median <- stats::median(placebo_gaps$mean_abs_gap)
detect_bound   <- stats::quantile(placebo_gaps$mean_abs_gap, 0.95)

money_2019 <- synth_effects(fits$money) |>
  dplyr::filter(time_unit == 2019) |>
  dplyr::pull(gap)
```

**Did Washington's McCleary response buy measurable achievement? No — and the
money did arrive, so that is a finding about the spending, not about the
data.**

Two results, and they point in different directions.

**The money is real and large.** By 2019 Washington was spending
**$`r formatC(round(money_2019), format = "d", big.mark = ",")` more per pupil**
than its synthetic counterpart, in inflation- and cost-of-living-adjusted
dollars — a divergence that ranks 4th of 40 units against placebo, with
pre-treatment tracking inside $35. The common objection that McCleary was
merely a levy swap, moving the same dollars from local to state ledgers, does
not survive the correction. Real resources per student rose, substantially,
relative to what would otherwise have happened.

**The achievement gain is absent.** Washington's grade 8 math scores diverge
from synthetic Washington by an average of
**`r round(wa_abs_gap, 2)` scale points** across the post-treatment period. The
median untreated state diverges by `r round(placebo_median, 2)` points, which
is the same number to within a rounding error. Its post/pre RMSPE ratio ranks
`r inf$rank` of `r nrow(inf$table)`
(*p* = `r round(inf$p_value, 3)`), and the result holds whether treatment is
dated 2013, 2015, or 2017, and whether or not the donor exclusions are applied.

### How large an effect could have hidden?

This is the question that separates "no effect" from "no power to see one," and
it deserves a number rather than a caveat.

A sustained divergence would have to average about
**`r round(detect_bound, 1)` scale points** to fall outside the 95th percentile
of what untreated states produce by chance in this design. That is the
detection floor. For scale: Washington's entire lead over the national public
average was 5.0 points in 2003 and 1.4 points in 2024.

So the honest statement has two halves, and both are firm:

- **An effect large enough to matter in the way McCleary's advocates argued —
  one that would have visibly moved Washington's standing against the nation —
  is ruled out.** It is not in the data. The post-treatment gaps (+3.9, +2.6,
  +4.1, +0.9 through 2019) are the wrong size *and* well inside the placebo
  band.
- **A modest effect — one or two scale points, sustained — cannot be ruled out
  and never could have been**, with five pre-treatment periods and a biennial
  test. If that is the size of the true effect, this design was never going to
  find it, and neither will any other design built on state-level NAEP.

Undetectable at this resolution is not the same as zero, and a reader hoping
to find McCleary vindicated here will not find that either, because there is no
positive result to point at.

::: {.callout-important}
## Three things that constrain the verdict

**The donor pool is concentrated.** Synthetic Washington is
`r scales::percent(max(tidysynth::grab_unit_weights(fits$score)$weight), accuracy = 0.1)`
South Dakota, with 90% of the weight in three states. By the standard set in
`notes/methodology.qmd`, that makes the point estimate illustrative rather than
inferential — the *direction* of the finding is robust across specifications,
but the precise gap in any given year is not.

**2022 and 2024 are pandemic-contaminated.** Both Washington and its synthetic
fall sharply; the negative gaps in those years should not be read as McCleary
failing. 2019 is the last clean post-treatment observation, and it is also the
year the money peaked.

**The test's power comes from the placebo distribution, not from theory.** The
5.5-point floor above is an empirical property of this donor pool and this
panel length, not a general statement about synthetic control.
:::

## What this can and cannot settle

**It does** settle the narrow question it was built for: after a large,
court-forced, substantially uniform resource increase, Washington's grade 8
mathematics performance did not diverge from comparable states by any amount
this design can see. The treatment was delivered and the outcome did not move.

**It cannot** tell you whether money works in general. The best-identified
evidence that school spending raises achievement comes from reforms that
targeted low-income districts [@jackson2016effects; @lafortune2018school].
McCleary was substantially uniform, and the accompanying levy cap partly
redistributed *away from* high-levy districts. A null for McCleary is
consistent with progressive finance reform working elsewhere.

**It cannot** evaluate what Bazzaz identifies as Washington's actual weak
spots — preschool access and postsecondary completion
[@bazzaz2026paramount], neither of which the constitutional guarantee covers.
Those need a different design, and the outcome data for them is worse. One
reading of all this is that the paramount-duty clause did exactly what it was
interpreted to require, funding K–12 to the point where additional dollars
there buy little, while the parts of a child's education with the weakest
measured results are the parts the clause never reached.

## Reproducing

```bash
make docker-build   # pinned R + Quarto image
make all            # fetch -> test -> render
```

See `notes/methodology.qmd` for specification checks and `README.md` for the
data-refresh cadence.
