---
title: "What Did McCleary Buy?"
subtitle: "A synthetic-control estimate of Washington's school finance reform"
author: "Richard Sprague"
date: today
---

```{r setup}
#| include: false
source(here::here("R", "00_config.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "04_build_panel.R"))
source(here::here("R", "05_synth.R"))
source(here::here("R", "06_plots.R"))

panel <- build_analysis_panel()
fits  <- run_all(panel)
inf   <- synth_inference(fits$score)
```

## The question

Between the 2009–11 and 2019–21 biennia, Washington's state spending on K–12
education rose 110%, while all other state spending rose 52% [@wrc2020mccleary].
Over the same period the state's standing on the National Assessment of
Educational Progress did not improve relative to the rest of the country. In
grade 8 mathematics Washington led the national public average by 5 points in
2003 and by 2 points in 2024.

Those two facts are usually deployed as a rhetorical pair. They should not be.
Three things have to be settled before they can be compared at all:

1. **How much money actually reached schools per student?** The 110% figure
   measures the state budget line. McCleary's central mechanism was a levy
   swap — the state assumed obligations previously met by local levies — so a
   large share of that growth is a change in *who writes the check*, not in how
   much is written. Total revenue per pupil rose 49.7% over 2010–11 to 2018–19,
   with state sources up 81.6% and local sources down 0.3%.
2. **What were those dollars worth?** Nominal per-pupil growth has to be
   deflated twice: once for inflation, once for the fact that Puget Sound labor
   got more expensive relative to the rest of the country over the same decade.
3. **Compared to what?** Washington's post-2019 scores fell. So did everyone's.
   The counterfactual is not "Washington in 2011"; it is "Washington had the
   legislature not responded to McCleary."

This project answers the third question with a synthetic control, and carries
the first two through the data pipeline so the resource measure being tested is
one that a skeptic would accept.

## Data

```{r panel-summary}
#| label: tbl-panel
#| tbl-cap: "Analysis panel coverage"
panel |>
  dplyr::summarise(
    States = dplyr::n_distinct(state),
    `NAEP years` = dplyr::n_distinct(year),
    `First year` = min(year),
    `Last year` = max(year),
    Observations = dplyr::n()
  )
```

Three sources, all public, all pulled by script:

| Source | Series | Access |
|---|---|---|
| NAEP Data Service (NCES) | State mean scale scores, grade 8 math | `R/01_fetch_naep.R` |
| Census F-33 / CCD via Urban Institute | Revenue and expenditure per pupil | `R/02_fetch_finance.R` |
| BEA Regional Price Parities; CPI-U | Place and time deflators | `R/03_price_adjust.R` |

Every API pull is validated against hand-transcribed published figures in
`data/cache/naep_verified_anchors.csv` before the analysis runs
(`make test`). This guards against the failure mode where an upstream schema
change silently returns a different series and the analysis keeps running.

### The spending series, deflated

```{r fig-spending}
#| label: fig-spending
#| fig-cap: "Washington per-pupil revenue, nominal and deflated"
plot_spending(panel)
```

```{r tbl-spending}
#| label: tbl-spending
#| tbl-cap: "Washington per-pupil revenue by NAEP year"
panel |>
  dplyr::filter(state == TREAT_UNIT) |>
  dplyr::transmute(
    Year = year,
    `Nominal $/pupil` = round(rev_pp),
    `Real, cost-adj. $/pupil` = round(rev_pp_real),
    `State share of revenue` = scales::percent(state_share, accuracy = 0.1)
  )
```

## Design

Synthetic control [@abadie2010synthetic] constructs a weighted combination of
untreated states that tracks Washington's pre-2013 outcome path, then asks
whether Washington diverges from that combination afterward.

Treatment is dated to **2013**, the first NAEP administration after the January
2012 ruling. The money phased in through 2019, so the estimated effect is a
dose that grows over the post-period rather than a clean step.

The donor pool excludes states with their own major finance-reform shocks in
2008–2022 — `r paste(EXCLUDE_DONORS, collapse = ", ")` — which leaves
`r length(donor_states(panel))` donors. That exclusion list is the most
contestable input in this analysis; `notes/methodology.qmd` reports how much it
matters.

## Results

### First stage: did the money actually move?

A finance reform that does not change real resources cannot be tested for
effects on achievement. This is the check that the treatment exists.

```{r fig-first-stage}
#| label: fig-first-stage
#| fig-cap: "Real cost-adjusted revenue per pupil: Washington vs synthetic Washington"
plot_trends(
  fits$money,
  title = "Washington did buy something",
  subtitle = "Real, cost-of-living-adjusted revenue per pupil against a synthetic control built from untreated states",
  y_lab = "Revenue per pupil (2019 $, cost-adjusted)"
)
```

### Outcome: NAEP grade 8 mathematics

```{r fig-outcome}
#| label: fig-outcome
#| fig-cap: "NAEP grade 8 math: Washington vs synthetic Washington"
plot_trends(
  fits$score,
  title = "Observed and synthetic Washington",
  subtitle = "Grade 8 mathematics mean scale score"
)
```

```{r fig-gap}
#| label: fig-gap
#| fig-cap: "Estimated effect: observed minus synthetic"
plot_gap(
  fits$score,
  title = "Estimated effect of the McCleary response on grade 8 math",
  subtitle = "Positive values mean Washington outperformed its synthetic counterfactual"
)
```

### Inference

```{r fig-placebo}
#| label: fig-placebo
#| fig-cap: "Placebo distribution"
plot_placebos(
  fits$score,
  title = "Washington's gap against every donor state's placebo gap",
  subtitle = "If Washington's line is unremarkable inside the gray band, the estimate is indistinguishable from noise"
)
```

```{r tbl-inference}
#| label: tbl-inference
#| tbl-cap: "Post/pre RMSPE ratios, ranked"
inf$table |> head(12)
```

Washington's post/pre RMSPE ratio is `r round(inf$rmspe_ratio, 2)`, ranking
`r inf$rank` of `r nrow(inf$table)` units, giving a placebo-based
*p* = `r round(inf$p_value, 3)`.

::: {.callout-important}
## Read the inference before the point estimate

With five pre-treatment periods, the donor pool can fit almost any path, and
the placebo test has little power. A null here is weak evidence of no effect,
not evidence of a null effect. The 2022 and 2024 observations are also
contaminated by pandemic disruption that varied across states in ways the
pre-2013 fit cannot absorb. Treat the 2019 observation as the only clean
post-treatment point, and treat the whole exercise as a bound on how large an
effect could have gone undetected.
:::

## What this can and cannot settle

**It can** establish whether Washington's achievement diverged from comparable
states after a large, court-forced, mostly non-progressive resource increase —
and how large an effect the data could have hidden.

**It cannot** tell you whether money works in general. The best-identified
evidence that school spending raises achievement comes from reforms that
targeted low-income districts [@jackson2016effects; @lafortune2018school].
McCleary was substantially uniform, and the accompanying levy cap partly
redistributed *away from* high-levy districts. A null for McCleary is
consistent with progressive finance reform working elsewhere.

**It cannot** evaluate the article's actual claim — that Washington's weak
spots are pre-K access and college completion, neither of which the
constitutional guarantee covers. That is a separate design, and the outcome
data for it is worse.

## Reproducing

```bash
make docker-build   # pinned R + Quarto image
make all            # fetch -> test -> render
```

See `notes/methodology.qmd` for specification checks and `README.md` for the
data-refresh cadence.
