---
title: "Methodology and specification checks"
---

```{r setup}
#| include: false
source(here::here("R", "00_config.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "04_build_panel.R"))
source(here::here("R", "05_synth.R"))
source(here::here("R", "06_plots.R"))
panel <- build_analysis_panel()
```

## Decisions a reviewer should attack

Every one of these is a knob in `R/00_config.R`.

### 1. Treatment date

`TREAT_YEAR <- 2013`. Defensible alternatives: 2012 (ruling), 2015 (first
substantial appropriations), 2019 (full phase-in). The later the date, the fewer
post-periods remain and the more the pandemic dominates them.

```{r alt-treat}
#| label: tbl-alt-treat
#| tbl-cap: "Sensitivity to treatment date"
purrr::map_dfr(c(2013, 2015, 2017), function(ty) {
  sc <- fit_synth(panel, "score", treat_year = ty)
  s  <- synth_inference(sc)
  tibble::tibble(
    `Treatment year` = ty,
    `RMSPE ratio` = round(s$rmspe_ratio, 2),
    `Rank` = s$rank,
    `p` = round(s$p_value, 3)
  )
})
```

### 2. Donor pool exclusions

The exclusion list encodes a judgment about which states had their own
concurrent finance shocks. Here is the same estimate with no exclusions at all.

```{r alt-donors}
#| label: tbl-alt-donors
#| tbl-cap: "Sensitivity to donor-pool exclusions"
all_donors <- setdiff(unique(panel$state), c(TREAT_UNIT, NON_STATES))
sc_all <- fit_synth(panel, "score", donors = all_donors)
s_all  <- synth_inference(sc_all)

tibble::tibble(
  Specification = c("Excluding reform states", "All states as donors"),
  `Donors` = c(length(donor_states(panel)), length(all_donors)),
  `RMSPE ratio` = c(round(synth_inference(fit_synth(panel, "score"))$rmspe_ratio, 2),
                    round(s_all$rmspe_ratio, 2)),
  `p` = c(round(synth_inference(fit_synth(panel, "score"))$p_value, 3),
          round(s_all$p_value, 3))
)
```

### 3. Pre-period length

`PRE_START <- 2003` buys a balanced panel at the cost of five pre-periods.
Setting it to 1996 adds pre-periods but drops states that did not participate in
state NAEP before NCLB. Rerun with `PRE_START <- 1996` and compare donor counts.

### 4. The place deflator

BEA Regional Price Parities measure the cost of a consumption basket. What a
school district actually faces is the *wage* of college-educated labor in its
market, which is what NCES's Comparable Wage Index for Teachers (CWIFT)
measures. CWIFT is the better instrument and is not API-served; drop
`data/raw/cwift.csv` in place and `R/03_price_adjust.R` prefers it
automatically. Expect the adjustment to be larger under CWIFT than under RPP,
because Puget Sound tech wages pull teacher-comparable wages up faster than they
pull the consumption basket up.

### 5. Donor weights

```{r weights}
#| label: tbl-weights
#| tbl-cap: "Unit weights in the synthetic control"
fit_synth(panel, "score") |>
  tidysynth::grab_unit_weights() |>
  dplyr::filter(weight > 0.01) |>
  dplyr::arrange(dplyr::desc(weight)) |>
  dplyr::mutate(weight = round(weight, 3))
```

A synthetic control that leans on two or three donors is fragile. If the table
above is concentrated, report the estimate as illustrative rather than
inferential.

## Known threats to validity

- **Interference.** NAEP is a common instrument; there is no spillover from
  Washington's budget to other states. SUTVA is plausible here, unusually.
- **Anticipation.** Districts may have adjusted before 2013 in expectation of
  the ruling, which would bias toward zero.
- **Compositional change.** Washington's English-learner and low-income shares
  moved over the window. NAEP scores are not demographically adjusted here.
  The Urban Institute publishes demographically adjusted state NAEP; swapping
  that in as the outcome is the single highest-value extension.
- **Pandemic confounding.** 2022 and 2024 reflect state-varying school closure
  duration, which correlates with politics and is not absorbed by pre-2013 fit.
  Washington closed longer than the median state, which would bias the estimate
  *downward* and is the strongest reason not to read the post-2020 gap as a
  McCleary effect.
