---
title: "Methodology and specification checks"
---

```{r setup}
#| include: false
source(here::here("R", "00_config.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "04_build_panel.R"))
source(here::here("R", "05_synth.R"))
source(here::here("R", "06_plots.R"))
panel <- build_analysis_panel()
```

## Decisions a reviewer should attack

Every one of these is a knob in `R/00_config.R`.

### 1. Treatment date

`TREAT_YEAR <- 2013`. Defensible alternatives: 2012 (ruling), 2015 (first
substantial appropriations), 2019 (full phase-in). The later the date, the fewer
post-periods remain and the more the pandemic dominates them.

```{r}
#| label: tbl-alt-treat
#| tbl-cap: "Sensitivity to treatment date"
purrr::map_dfr(c(2013, 2015, 2017), function(ty) {
  sc <- fit_synth(panel, "score", treat_year = ty)
  s  <- synth_inference(sc)
  tibble::tibble(
    `Treatment year` = ty,
    `RMSPE ratio` = round(s$rmspe_ratio, 2),
    `Rank` = s$rank,
    `p` = round(s$p_value, 3)
  )
})
```

### 2. Donor pool exclusions

The exclusion list encodes a judgment about which states had their own
concurrent finance shocks. Here is the same estimate with no exclusions at all.

```{r}
#| label: tbl-alt-donors
#| tbl-cap: "Sensitivity to donor-pool exclusions"
all_donors <- setdiff(unique(panel$state), c(TREAT_UNIT, NON_STATES))
sc_all <- fit_synth(panel, "score", donors = all_donors)
s_all  <- synth_inference(sc_all)

tibble::tibble(
  Specification = c("Excluding reform states", "All states as donors"),
  `Donors` = c(length(donor_states(panel)), length(all_donors)),
  `RMSPE ratio` = c(round(synth_inference(fit_synth(panel, "score"))$rmspe_ratio, 2),
                    round(s_all$rmspe_ratio, 2)),
  `p` = c(round(synth_inference(fit_synth(panel, "score"))$p_value, 3),
          round(s_all$p_value, 3))
)
```

### 3. Pre-period length

`PRE_START <- 2003` buys a balanced panel at the cost of five pre-periods.
Setting it to 1996 adds pre-periods but drops states that did not participate in
state NAEP before NCLB. Rerun with `PRE_START <- 1996` and compare donor counts.

### 4. The place deflator

BEA Regional Price Parities measure the cost of a consumption basket. What a
school district actually faces is the *wage* of college-educated labor in its
market, which is what NCES's Comparable Wage Index for Teachers (CWIFT)
measures. CWIFT is the better instrument and is not API-served; drop
`data/raw/cwift.csv` in place and `R/03_price_adjust.R` prefers it
automatically. Expect the adjustment to be larger under CWIFT than under RPP,
because Puget Sound tech wages pull teacher-comparable wages up faster than they
pull the consumption basket up.

```{r}
#| label: deflator-bound
#| include: false
fmt_usd <- function(x) paste0("$", formatC(round(x), format = "d", big.mark = ","))
post_gaps <- function(sc) {
  synth_effects(sc) |>
    dplyr::filter(time_unit >= TREAT_YEAR) |>
    dplyr::select(time_unit, gap) |>
    tibble::deframe()
}
fit_main    <- fit_money(panel)
fit_cpi     <- fit_money(panel, place_adjust = FALSE)
gap_main    <- post_gaps(fit_main)
gap_cpi     <- post_gaps(fit_cpi)
rank_main   <- synth_inference(fit_main)$rank
rank_cpi    <- synth_inference(fit_cpi)$rank
n_units     <- nrow(synth_inference(fit_main)$table)
place_name  <- if (file.exists(file.path(DIR_RAW, "cwift.csv"))) "CWIFT" else "RPP"
wa_place    <- panel |>
  dplyr::filter(state == TREAT_UNIT) |>
  dplyr::distinct(year, place_index) |>
  tibble::deframe()
```

The place deflator also cuts against the treatment itself. Most of the
post-McCleary money went to teacher salaries. Deflating by a place index that
climbs with the regional labor market treats part of those raises as merely
keeping pace with local prices, which nets out some of what the state bought.
CWIFT, a wage index, has this problem most directly; RPP has a milder version
through housing costs. The main specification currently uses
`r place_name`, and Washington's index rose from
`r sprintf("%.2f", wa_place[["2011"]])` in 2011 to
`r sprintf("%.2f", wa_place[["2019"]])` in 2019. To bound the effect,
`fit_money(panel, place_adjust = FALSE)` in `R/05_synth.R` re-deflates revenue
by CPI-U alone. The 2019 first-stage gap
widens from `r fmt_usd(gap_main[["2019"]])` to `r fmt_usd(gap_cpi[["2019"]])`
(`r sprintf("%+.0f%%", 100 * (gap_cpi[["2019"]] / gap_main[["2019"]] - 1))`),
and Washington's placebo rank moves from `r rank_main` to `r rank_cpi` of
`r n_units`. The cost-adjusted figure reported on the main page is therefore
the conservative one, and switching to CWIFT would likely make it more
conservative still.

### 5. Donor weights

```{r}
#| label: tbl-weights
#| tbl-cap: "Unit weights in the synthetic control"
fit_synth(panel, "score") |>
  tidysynth::grab_unit_weights() |>
  dplyr::filter(weight > 0.01) |>
  dplyr::arrange(dplyr::desc(weight)) |>
  dplyr::mutate(weight = round(weight, 3))
```

A synthetic control that leans on two or three donors is fragile. If the table
above is concentrated, report the estimate as illustrative rather than
inferential.

## Known threats to validity

- **Interference.** NAEP is a common instrument; there is no spillover from
  Washington's budget to other states. SUTVA is plausible here, unusually.
- **Anticipation.** Districts may have adjusted before 2013 in expectation of
  the ruling, which would bias toward zero.
- **Compositional change.** Washington's English-learner and low-income shares
  moved over the window. NAEP scores are not demographically adjusted here.
  The Urban Institute publishes demographically adjusted state NAEP; swapping
  that in as the outcome is the single highest-value extension.
- **Pandemic confounding.** 2022 and 2024 reflect state-varying school closure
  duration, which correlates with politics and is not absorbed by pre-2013 fit.
  Washington closed longer than the median state, which would bias the estimate
  *downward* and is the strongest reason not to read the post-2020 gap as a
  McCleary effect.
